Increasing prevalence of plant-fungal symbiosis over two centuries of climate change

### 2023-03-23 Manuscript/endo_herbarium.tex
edits to the discussion and lots of stuff

\documentclass[11pt]{article}
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}

@@

\title{Increasing prevalence of plant-fungal symbiosis over 195 years of climate change}

@@

Understanding how biotic interactions are altered by global climate change is a major goal of basic ecological research and conservation management \cite{gilman2010framework,blois2013climate}.
Documented responses to environmental change, such as shifts in species' distributions \cite{aitken2008adaptation} and phenology \cite{piao2019plant}, are typically blind to concurrent changes in associated biotic interactions.
Empirically evaluating these biotic changes, whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will}, is crucial to predicting future climate change responses, but has been limited because, simply put, data on species interactions are limited, with few datasets extending over the long time scales of contemporary climate change \cite{poisot2021global}.

@@

\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}
There are transmitted vertically from maternal plants to offspring through the grasses' seeds.
Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
Over time, endophytes that act as mutualists for their hosts should rise in prevalence within a population. 
\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic}, and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
Previous survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
We predicted that prevalence should also track temporal changes in environmental drivers that elicit these fitness benefits.
For example, endophyte-mediated drought tolerance should lead prevalence to increase in regions where precipitation declines over time.
Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}. 
Grasses are commonly identified and collected based on the presence of their reproductive structures meaning that collected specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets. 
This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change.

In this study, we assessed for the first time the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three \emph{Epichloë} endophyte host species, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How does endophyte prevalence vary across space?, (ii) How has endophyte prevalence changed over time? and (iii) How have temporal changes in endophyte prevalence differed across space?
We evaluated the predictive ability of this climate-implicit model using contemporary endophyte surveys as "out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical}. 
We then address how climate change may be driving these trends in endophyte prevalence by addressing the question: (iv) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
We examined 1,865 specimens collected across across eastern North America between 1824 and 2019.

\section*{Methods}
        \subsection*{Focal species}
Our surveys focused on three endophyte hosts, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the Eastern United States.
\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically blooms between March and July (most common collection month: May).
\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and blooms in late Summer and early Autumn (most common collection month: September). 
This species is relatively more sparsely distributed, tending to be found in shadier and more moist habitats while \emph{A. hyemalis} is often found in open and recently disturbed ground. 
Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A. perennans} has a slighty more northern distribution, whereas \emph{A. hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed that the \emph{Agrostis} species. 
It begins flowering as early as March or April but continues throughout the Summer (most common collection month: July).

		\subsection*{Herbarium surveys}
We visited herbaria between 2019 and 2022, including The Botanical Research Institude of Texas (which also houses the Vanderbilt University herbarium and the R. Dale Thomas Plant collection from the University of Louisiana at Monroe), the Louisiana State University Herbarium, the Mercer Botanic Garden Herbarium, Missouri Botanic Garden's Herbarium, the S.M. Tracy Herbarium at Texas A\&M, the University of Kansas Herbarium, the Robert Bebb Herbarium at the University of Oklahoma, the Lundell Herbarium and the UT Herbarium at the University of Texas, and Oklahoma State University Herbarium (See Table A1 for a summary of specimens includes from each collection).
Our sampling plan was designed to minimize damage to these specimens.
We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies. 
We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected viable seeds fit for our purpose of identifying fungal endophytes.

After collecting seeds, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, using microscopy. 
We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue dye and squashed them under a microscope cover slip. 
We examined the squashed seeds for the presence of fungal hyphae at 100X magnification \cite{bacon2018stains}.
In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded from the seed counts for that specimen.
On average we scored 4.7 seeds per specimen in \emph{A. hyemalis}, 5 seeds per specimen in \emph{A. perennans}, and 3.8 seeds per specimen in \emph{E. virginicus}.
Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
We therefore designated each specimen as symbiotic if endophyte hyphae were observed in one or more seeds, or non-symbiotic if hyphae were observed in zero seeds. 
To capture uncertainty in the identification process, we recorded both a "liberal" and a "conservative" endophyte status for each plant.  
When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be a true endophyte) and a negative conservative status (less likely to be a true endophyte). 
$89$\% of scored plants had matching liberal and conservative endophyte statuses, reflecting high confidence in endophyte identification.
The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Figure A5). 
We excluded specimens for which information about the collection location and date were unavailable.
In total, we quantified endophyte symbiosis for $964$ \emph{A. hyemalis} specimens collected between 1824 and 2019, $299$ \emph{A. perennans} specimens collected between 1863 and 2017, and $602$ \emph{E. virginicus} specimens collected between 1839 and 2017 (Fig. 1, Fig 2A).
Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
Many specimens had digitized collection information readily available, but for those that did not, we transcribed information from pictures of the specimens. 
Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid.

@@

The expected endophyte prevalence, $\hat{P}$, was modelled with intercepts specific to each host species ($\beta_{0}$), slopes for changes over time ($\beta_{1}$) as well as the interaction between time and the specimen's latitude and longitude ($\beta_{2}$, $\beta_{3}$, and $\beta_{4}$). 
To quantify spatial patterns in prevalence, we included a spatially-dependent random intercept, $\phi$.
This random spatial effect was modeled as a stochastic partial differential equation (SPDE) that depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
The covariance matrix is approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and to each scorer $\omega$.
We performed modeling fitting using the rINLA package, \citep{lindgren2015bayesian}, with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
Posterior modes were stable indicating that numeric convergence was successful.
We assessed model fit with graphical posterior predictive checks (Fig. A3).
The model performed adequately at classifying the historical data (AUC = 0.77; Fig. A4).

We used data from contemporary surveys of on endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} as test data to evaluate predictions of the model. 
During these surveys, which took place between 2013 and 2020, we collected seeds from living plants from 63 populations(average number of plants sampled: 22.9) across a portion of each species' range (Fig A4). 
We quantified the endophyte status with staining microscopy as described for the herbarium surveys and calculated prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
We compared the model's prediction for these locations to the observed population prevalence.

@@

We calculated the relative change in endophyte prevalence across the 1925 to 2020, i.e., the difference between predicted prevalence divided by the endophyte prevalence in 1925, because regions varied in their predicted starting prevalence. 
This time period connects the predicted change in prevalence with the endpoints of the available climate record.
We then calculated the Pearson correlation coefficient between the relative change in endophyte prevalence and each climate driver.

@@

\section*{Results}
\subsection*{Temporal and spatial trends}
We quantified temporal and spatial trends in endophyte prevalence while accounting for potential biases introduced by collectors and by individuals who quantified endophyte presence/absence with the use of random effects. We found no evidence that collector biases influenced our results (random effects values centered around 0; Fig 2A), while the identity of individual scorers did contribute to observed patterns in endophyte prevalence (3 of the 16 scorers were more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals that do not overlap 0) (Fig 2B). Interpretation of the data without accounting for this source of variation would lead to potentially biased interpretations of temporal and spatial trends.

\begin{figure}[H]
	\label{fig:random_fx}
	\centering
	\includegraphics[width = .8\linewidth]{random_fx_plot.png}
	\caption{\textbf{Posterior estimates of (A) collector and (B) scorer random effects.} Points show tthe posterior mean along with 95\% CI for random effects estimate from 532 collectors and 16 scorers.}
\end{figure}

\subsubsection*{How does endophyte prevalence vary across space?}
Across space, there were clear trends in endophyte prevalence which varied between species (Fig. 3).
\emph{Elymus virginicus} had higher prevalence towards the northern portions of its range. 
In contrast, \emph{Agrostis hyemalis} had highest prevalence towards the center of its range with regions of low prevalence towards the northeast and towards its western range edge.
\emph{Agrostis perennans} had  highest prevalence towards the wester portion portion of it's range edge.
There is considerable uncertainty in the spatial pattern of endophyte prevalence (See Fig. A10 for projections of the 95\% credible interval), however these broad spatial patterns are consistent across the low and high probability model predictions.

@@

\subsubsection*{How have temporal changes in endophyte prevalence differed across space?}
Our model revealed that while there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges.
Temporal trends vary with latitude and longitude, as evidenced by variation in the spatial pattern of endophyte prevalence through time in Fig. 3, and that temporal trends in Fig. 4 vary at different rates according to their latitude.
In some regions, \emph{A. perennans} experienced increases in percent prevalence by as much as 35 percentage points between 1925 and 2020, while \emph{A. hyemalis} and \emph{E. virginicus} experienced increases up to around 15 percentage points. 
In other regions, there were smaller increases of only five percentage points for \emph{E. virginicus} and fifteen percentage points for \emph{A. perennans}. 
We calculated relative change in endophyte prevalence because regions differ in their initial endophyte prevalence.

@@

This reveals locations that are hotspots of change in endophyte prevalence (Fig. 5). 
The overall magnitude of changes were greatest in \emph{A. perennans}, particularly in the northeast and in the south, where endophytes were previously only found in low prevalence (Fig. 3).
\emph{A. hyemalis} and \emph{E. virginicus} experience lower overall increases in prevalence, yet we can still identify hotspots of change in the northeast for \emph{A. hyemalis} and in the southeast for \emph{E. virginicus}.

@@

We found that trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. 6).
For the majority of the study region, the climate has become wetter and cooler over the last century (Fig. A5), a consequence of regional variation in global climate change \cite{ipcc_2021}. 
Within the study region, spatially heterogeneous environmental changes were predictive of changes in endophyte prevalence. 
For example, strong increases in prevalence within \emph{E. virginicus} were most associated (as indicated by correlation coefficients $>.3$) with declines in Summer precipitation  (a negative correlation in Fig. 4) as well as with increases in the year-to-year variability of Summer precipitation (a positive correlation in Fig. 4). 
Changes were also associated with reductions in average annual temperatures, and increases in year-to-year temperature variability.
\emph{A. perennans} endophyte prevalence increased most strongly in regions that experienced reduced spring precipitation and reduced variability in annual temperature.
Although these correlations were weaker, changes in \emph{A. perennans} endophyte prevalence were also associated with increased in increases in annual precipitation and increasing autumn precipitation. 
For \emph{A. hyemalis}, endophyte prevalence increased most strongly in regions that experienced reductions in autumn precipitation variability.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{climate_corr_heatmap.png}
	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Color denotes the Pearson correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver. Positive correlation coefficients indicate that greater increases in a climate driver were associated with larger increases in endophyte prevalence, while negative values indicate that . Asteriks denote correlation coefficients $\> .3$ or $< -.3$.}
\end{figure}

\section*{Discussion}
Our examination of historic plant specimens revealed a cryptic biotic reponse to climate change. 
For the three host species we examined, there have been clear increases in fungal endophyte prevalence over the last two centuries.
This increase suggests that \emph{Epichloë} fungal endophytes are playing key \tom{adaptive}{In what way are they adaptive?} roles in response to \tom{climate change}{Climate change is not the only change over this time period, so be careful with attribution.} and that the mutualism between grasses and their fungal endophytes has contributed to resilience for each partner species under the observed changes in climate. 
Spatial and temporal trends in endophyte prevalence varied between each species in \tom{ecologically meaningful ways}{vague}.
\tom{Our spatially-explicit model predicted regions of both high and low endophyte prevalence, suggesting that symbiotic and non-symbiotic host plants have overlapping, but non-identical niche requirements.}{This is a space-only result and interpretation, which should be better explained and motivated.} 
\tom{Endophytes may expand the range of their hosts by allowing them to persist in environments where they otherwise could not \citep{afkhami2014mutualist, kazenel2015mutualistic}.}{Not sure this is right for the first paragraph.}
Understanding how microbial symbionts respond to climate change is a first step towards incorporating reciprocal host-symbiont interactions into predictions host range shifts.
\tom{}{This paragraph should address the core question of whether climate change strengthens or disrupts mutualism.}

>>>>

\documentclass[11pt]{article}
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,round,sectionbib,sort]{natbib}   % omit 'round' option if you prefer square brackets
\let\cite\citep

\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}

@@

\title{Increasing prevalence of plant-fungal symbiosis over two centuries of climate change}

@@

Understanding how biotic interactions are altered by global climate change is a major goal of basic ecological research and conservation management \cite{gilman2010framework,blois2013climate}.
Documented responses to environmental change, such as shifts in species' distributions \cite{aitken2008adaptation} and phenology \cite{piao2019plant}, are typically blind to concurrent changes in associated biotic interactions.
Empirically evaluating these biotic changes, whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will}, is crucial to predicting future climate change responses. 
Such evaluations have been limited because data on species interactions are limited, with few datasets extending over the long time scales of contemporary climate change \cite{poisot2021global}.

@@

\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}
They are transmitted vertically from maternal plants to offspring through the grasses' seeds.
Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
Over time, endophytes that act as mutualists for their hosts should rise in prevalence within a population. 
\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic}, and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
Previous survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
We predicted that prevalence should also track temporal changes in environmental drivers that elicit these fitness benefits.
For example, endophyte-mediated drought tolerance should lead prevalence to increase in regions where precipitation declines over time.
Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}. 
Grasses are commonly identified and collected based on the presence of their reproductive structures meaning that collected specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets. 
This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change.

In this study, we assessed for the first time the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three \emph{Epichloë} endophyte host species, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How does endophyte prevalence vary across space?, (ii) How has endophyte prevalence changed over time? and (iii) How have temporal changes in endophyte prevalence differed across space?
To answer these questions, we fit a spatio-temporal statistical model of endophyte prevalence.
and evaluated the predictive ability of the model using contemporary endophyte surveys as out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical}. 
We then address how climate change may be driving these trends in endophyte prevalence by addressing the question: (iv) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
We examined 1,865 specimens collected across across eastern North America between 1824 and 2019.

\section*{Methods}
        \subsection*{Focal species}
Our surveys focused on three endophyte hosts, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the Eastern United States.
\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically blooms between March and July (most common collection month: May).
\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and blooms in late Summer and early Autumn (most common collection month: September). 
This species is relatively more sparsely distributed, tending to be found in shadier and more moist habitats while \emph{A. hyemalis} is often found in open and recently disturbed ground. 
Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A. perennans} has a slighty more northern distribution, whereas \emph{A. hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed than the \emph{Agrostis} species. 
It begins flowering as early as March or April but continues throughout the Summer (most common collection month: July).

		\subsection*{Herbarium surveys}
We visited herbaria between 2019 and 2022, including The Botanical Research Institude of Texas (which also houses the Vanderbilt University herbarium and the R. Dale Thomas Plant collection from the University of Louisiana at Monroe), the Louisiana State University Herbarium, the Mercer Botanic Garden Herbarium, Missouri Botanic Garden's Herbarium, the S.M. Tracy Herbarium at Texas A\&M, the University of Kansas Herbarium, the Robert Bebb Herbarium at the University of Oklahoma, the Lundell Herbarium and the UT Herbarium at the University of Texas, and Oklahoma State University Herbarium (See Table A1 for a summary of specimens included from each collection). 
In total, we quantified endophyte symbiosis for $964$ \emph{A. hyemalis} specimens collected between 1824 and 2019, $299$ \emph{A. perennans} specimens collected between 1863 and 2017, and $602$ \emph{E. virginicus} specimens collected between 1839 and 2017 (Fig. 1, Fig 2A).
Our sampling plan was designed to minimize damage to these specimens.
We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies. 
We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
We excluded specimens for which information about the collection location and date were unavailable.
Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
Many specimens had digitized collection information readily available, but for those that did not, we transcribed information from pictures of the specimens. 
Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid.

After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, in each specimen using microscopy. 
We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue dye and squashed them under a microscope cover slip. 
We examined the squashed seeds for the presence of fungal hyphae at 100X magnification \cite{bacon2018stains}.
In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded from the seed counts for that specimen.
On average we scored 4.7 seeds per specimen in \emph{A. hyemalis}, 5 seeds per specimen in \emph{A. perennans}, and 3.8 seeds per specimen in \emph{E. virginicus}.
Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
We therefore designated each specimen as symbiotic if endophyte hyphae were observed in one or more seeds, or non-symbiotic if hyphae were observed in zero seeds. 
To capture uncertainty in the identification process, we recorded both a "liberal" and a "conservative" endophyte status for each plant.  
When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be a true endophyte) and a negative conservative status (less likely to be a true endophyte). 
$89$\% of scored plants had matching liberal and conservative endophyte statuses, reflecting high confidence in endophyte identification.
The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Figure A5).

@@

The expected endophyte prevalence, $\hat{P}$, was modelled with intercepts specific to each host species ($\beta_{0}$), slopes for changes over time ($\beta_{1}$) as well as the interaction between time and the specimen's latitude and longitude ($\beta_{2}$, $\beta_{3}$, and $\beta_{4}$). 
To quantify spatial patterns in prevalence, we included a spatially-dependent random intercept, $\phi$.
This random spatial effect was modeled as a stochastic partial differential equation (SPDE) that depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
The SPDE model allowed us to flexibly estimate endophyte prevalence at locations across the study region, while accounting for potential spatial autocorrelation between data points via the covariance matrix. 
The covariance matrix was approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and to each scorer $\omega$.
Previous work has questioned whether the behavior of historical botanists and uneven sampling may introduce biases into ecological inferences made from historic collections \cite{kozlov2020biases}. 
Prolific collectors who contribute thousands of specimens may be more or less likely to collect certain species, or specimens with certain traits \cite{daru2018widespread}. 
Similarly, the process of scoring seeds for hyphae involved many individual researchers, who may vary in their likelihood of positively identifying fungal hyphae. 
By including a random effect for collectors and for scorers, we can account for the variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
We performed model fitting using the rINLA package, \citep{lindgren2015bayesian}, with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
Posterior modes were stable indicating that numeric convergence was successful.
We assessed model fit with graphical posterior predictive checks (Fig. A3).
The model performed adequately at classifying the historical data, comparing the accuracy of predictions from the model with observed data (AUC = 0.77; Fig. A4). 
\subsection*{Validating the model with an out-of-sample test}
We used data from contemporary surveys of endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} as test data to evaluate predictions of the model. 
Surveys of \emph{E. virginicus} were conducted in 2013 as described in \citet{sneck2017variation}, and surveys of \emph{A. hyemalis} took place between 2015 and 2020.
Population surveys of \emph{A. hyemalis} were initially designed to cover longitudinal variation in endophyte prevalence towards its range edge, while surveys of \emph{E. virginicus} were designed to cover latitudinal variation along its range edge. 
In total, we visited 43 populations of \emph{A. hyemalis} and 20 populations of \emph{E. virginicus} across the central southeastern US, in Texas and neighboring states (Fig A4).
During surveys, we collected seeds from up to 30 individuals per location  (average number of plants sampled: 22.9).
We quantified the endophyte status of each individual with staining microscopy as described for the herbarium surveys, and calculated the prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
The contempory survey period (2013-2020) is at the most recent edge of the time period encompassed by the historical observations used for model fitting.
We compared the model's prediction for these locations to the observed population prevalence.

@@

We calculated the relative change in endophyte prevalence across the 1925 to 2020, i.e., the difference between predicted prevalence divided by the endophyte prevalence in 1925, because regions varied in their predicted starting prevalence. 
This time period connects the predicted change in prevalence with the endpoints of the available climate record.
We then calculated the Spearman correlation coefficient between the relative change in endophyte prevalence and each climate driver.
Calculating correlations from many pixels across the study region risks articially inflating confidence in our results due to large sample sizes, and so we repeated this calculation using only a random subsample of 100 pixels across the study region.

@@

\section*{Results}
\subsection*{Temporal and spatial trends}
We quantified temporal and spatial trends in endophyte prevalence while accounting for potential biases introduced by collectors and by individuals who quantified endophyte presence/absence with the use of random effects. We found no evidence that collector biases influenced our results (collector random effects were consistently small; Fig 2A), while the identity of individual scorers did contribute to observed patterns in endophyte prevalence (3 of the 16 scorers were more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals that do not overlap 0) (Fig 2B). Interpretation of the data without accounting for this source of variation would lead to potentially biased interpretations of temporal and spatial trends.

\begin{figure}[H]
	\label{fig:random_fx}
	\centering
	\includegraphics[width = .8\linewidth]{random_fx_plot.png}
	\caption{\textbf{Posterior estimates of (A) collector and (B) scorer random effects.} Points show the posterior mean along with 95\% CI for random effects estimate from 532 collectors and 16 scorers.}
\end{figure}

\subsubsection*{How does endophyte prevalence vary across space?}
Across space, there were clear trends in endophyte prevalence which varied between species (Fig. 3).
We mapped predictions of endophyte prevalence across the area encompassing the sampled herbarium specimens for each species, and area smaller than the full geographic distribution of each species in nature.
\emph{Elymus virginicus} had higher prevalence towards the northern portions of the collection range. 
In contrast, \emph{Agrostis hyemalis} had highest prevalence towards the center of its collection range with regions of low prevalence towards the northeast and towards its western range edge.
\emph{Agrostis perennans} had  highest prevalence towards the westernmost portion of its collection range.
There is considerable uncertainty in the spatial pattern of endophyte prevalence (See Fig. A10 for projections of the 95\% credible interval), however these broad spatial patterns are consistent across the low and high end of model predictions.

@@

\subsubsection*{How have temporal changes in endophyte prevalence differed across space?}
Our model revealed that while there was an overall increase in endophyte prevalence, these changes varied across the host species' ranges.
The posterior estimates of the parameters describing space by time interactions, $\beta2$, $\beta3$, and $\beta4$ for each species indicate small effect sizes, but high certainty(i.e, a 99\% posterior probability of a negative interaction between year and latitude for \emph{A. hyemalis}, or a 67\% postterior probabilty of a postitive interaction between year and latitude for \emph{E. virginicus}) (Fig. A5).
In some regions, \emph{A. perennans} experienced increases in percent prevalence by as much as 35 percentage points between 1925 and 2020, while \emph{A. hyemalis} and \emph{E. virginicus} experienced increases up to around 15 percentage points. 
In other regions, there were smaller increases of only five percentage points for \emph{E. virginicus} and fifteen percentage points for \emph{A. perennans}. 
We calculated relative change in endophyte prevalence because regions differ in their initial endophyte prevalence.

@@

This revealed locations that were hotspots of change in endophyte prevalence (Fig. 5). 
The overall magnitude of changes were greatest in \emph{A. perennans}, particularly in the northeast and in the south, where endophytes were previously only found at low prevalence (Fig. 3).
\emph{A. hyemalis} and \emph{E. virginicus} experience lower absolute increases in prevalence, yet we can still identify hotspots of change in the northeast for \emph{A. hyemalis} and in the southeast for \emph{E. virginicus}.

\subsubsection*{Performance on test data}
The model adequately predicted mean endophyte prevalence, but underpredicted variance in the contemporary test observations (Fig. 6 A).
The model was able to predict large scale spatial trends in endophyte prevalence, such as declining endophyte prevalence towards western longitudes in \emph{A. hyemalis} (Fig. 6 B-C).
We interpreted this to mean that the model captured regional spatial dynamics, but underpredicts local scale dynamics. We discuss potential model improvements in the Discussion.

@@

We found that trends in endophyte prevalence were strongly associated with seasonal climate change drivers (Fig. 6).
For the majority of the study region, the climate has become wetter and cooler over the last century (Fig. A7-A8), a consequence of regional variation in global climate change \cite{ipcc_2021}. 
Within the study region, spatially heterogeneous environmental changes were predictive of changes in endophyte prevalence. 
For example, strong increases in prevalence within \emph{E. virginicus} were most associated with declines in Summer precipitation  (a negative correlation in Fig. 7) as well as with increases in the year-to-year variability of annual temperature (a positive correlation in Fig. 7). 
Changes were also associated with reductions in average annual temperatures, and increases in year-to-year temperature variability.
\emph{A. perennans} endophyte prevalence increased most strongly in regions that experienced reduced spring precipitation and reduced variability in annual temperature.
Although these correlations were weaker, changes in \emph{A. perennans} endophyte prevalence were also associated with increased in increases in annual precipitation and increasing autumn precipitation. 
For \emph{A. hyemalis}, endophyte prevalence increased most strongly in regions that experienced reductions in autumn precipitation variability. 
Correlations using only a subsampling of pixels were qualitatively similar to these results (Fig. A11), suggesting that the patterns we find, particulary the strongest correlations are not spurious associations.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{climate_corr_heatmap.png}
	\caption{\textbf{Correlations between changes in climate drivers and changes in endophyte prevalence.} Color denotes the Spearman correlation coefficient between the relative rate of change in endophyte prevalence and the change in annual mean temperature ($^oC$) and total annual and seasonal precipitation (mm), as well as the change in the coefficient of variation of each climate driver. Positive correlation coefficients indicate that greater increases in a climate driver were associated with larger increases in endophyte prevalence, while negative values indicate that . Asteriks denote correlation coefficients $> .3$ or $< -.3$.}
\end{figure}

\section*{Discussion}
Our examination of historic plant specimens revealed a cryptic biotic reponse to climate change. 
For the three host species we examined, there have been clear increases in fungal endophyte prevalence over the last two centuries.
Increases in prevalence of \emph{Epichloë}, which are vertically transmitted, can potentially be interpreted as adaptive changes that improve the fitness of their hosts under stressful conditions.
This interpretation is in line with theory predicting that the positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors}.
We found that trends in endophyte prevalence varied across the collection range of each species in assocation with observed changes in climate drivers, suggesting that the endophytes have contributed to host resilience under environmental change.
Taken together, this suggests a strengthening of the mutualism over the last two centuries.
